---
title: "Hurricane Michael's Windfield"
author: "James B. Elsner"
date: October 25, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Load the data.
```{r}
L <- "michaeldata.csv"
MW.df <- read.csv(L, header = TRUE)
MW.df <- MW.df[-33, ] # duplicate location Ballard Parners and Challenger Learning Center remove one
names(MW.df)
dim(MW.df)
```

The data frame contains 47 wind speed measurements (one duplicate location, removed). Longitude and latitude coordinates of the sites are given along with the peak wind gust (`PeakGustSpeed3`). Create a spatial points data frame by specifying columns that contain the spatial coordinates. Then assign a geographic coordinate system and convert the wind speed to meters per second.

```{r}
library(sp)
MW.spdf <- MW.df
coordinates(MW.spdf) <- ~ Longitude + Latitude
ll <- "+proj=longlat +datum=NAD83"
proj4string(MW.spdf) <- CRS(ll)
MW.spdf$gust <- MW.spdf$PeakGustSpeed3 * .51444
summary(MW.spdf$gust)
```

Create a histogram.
```{r}
library(ggplot2)
ggplot(as.data.frame(MW.spdf), aes(gust)) +
  geom_histogram()
```

Map the data.
```{r}
library(tmap)
library(tmaptools)
library(dplyr)
library(albersusa)

states <- usa_sf("aeqd")
FLALGA <- states %>%
  filter(iso_3166_2 %in% c("FL", "GA", "AL"))
counties <- counties_sf("aeqd") %>%
  filter(iso_3166_2 %in% c("FL", "GA", "AL"))
Mitchell <- counties %>%
  filter(iso_3166_2 == "GA", name == "Mitchell")
Camilla <- geocode_OSM("Camilla, GA", as.sf = TRUE)

tm_shape(states) +
  tm_polygons() +
tm_shape(FLALGA, is.master = TRUE) +
  tm_polygons() +
  tm_text(text = "name", xmod = c(0, 0, 2)) +
tm_shape(Mitchell) +
  tm_polygons(col = "black") +
  tm_text(text = "name", xmod = 2.5, col = "black", size = .75) +
tm_shape(Camilla) +
  tm_dots(col = "red", size = .1) +
tm_scale_bar(position = c("left", "bottom")) +
tm_compass(position = c("left", "bottom")) +
tm_layout(bg.color="lightblue")


tm_shape(MW.spdf) +
  tm_bubbles("gust") +
tm_shape(state_sf) +
  tm_borders() +
tm_shape(counties_sf) +
  tm_borders(col = "grey")
```

Next we create a character string specifying the tags for a planar projection and transform the geographic coordinates of the site locations and map polygons to the projected coordinates. Here we use Albers equal-area projection with true scale at latitudes 23 and 30$^\circ$ N and include a tag to specify meters as the unit of distance.
```{r}
aea <- "+proj=aea +lat_1=23 +lat_2=30 +lat_0=26, +lon_0=-83 +units=m"
lcc <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m"
MW.spdf <- spTransform(MW.spdf, CRS(lcc))
```

### Empirical variogram

Trends can be included by specifying coordinate names (e.g., `~ lon + lat`).  Note that, although the data have a planar projection, the coordinate names are not changed from the original data set.
```{r}
library(gstat)
MW.v <- variogram(gust ~ Longitude + Latitude, 
                  data = MW.spdf)
```

To see the variogram values as a function of lag distance use the `plot` method on the variogram object. Or here you use the regular plot method and add text indicating the number of point pairs for each lag distance.
```{r}
plot(MW.v$dist/1000, MW.v$gamma, xlim = c(0, 130), ylim = c(0, 150),
   xlab = "Lagged distance (h) [km]",
   ylab = expression(paste("Semivariance (", gamma, ") [", m^2/s^2, "]")),
   las = 1, pch = 19)
grid()
points(MW.v$dist/1000, MW.v$gamma, pch = 19)
text(MW.v$dist/1000, MW.v$gamma, pos = 1, 
     labels = as.character(MW.v$np), cex = .5)
```

### Variogram model

Next we fit a model to the empirical variogram. The variogram model is a mathematical relationship defining the semivariance as a function of lag distance. We first save the family and the initial parameter guesses in a variogram model (`FR.vmi`) object by typing
```{r}
MW.vmi <- vgm(model = "Gau", psill = 70, 
              range = 60 * 1000, nugget = 0)
MW.vmi
```

The `psill` argument is the partial sill as the difference between the sill and the nugget. We get estimates of the parameter values from looking at the empirical variogram. 

Next we use the `fit.variogram()` function to improve the fit.  Given a set of initial parameter values, the method of weighted least squares improves the parameter estimates. Ordinary least squares is not appropriate as the semivariances are correlated across the lag distances and the precision on the estimates varies depending on the number of site pairs for a given lag.
```{r}
MW.vm <- fit.variogram(MW.v, MW.vmi)
MW.vm
```

The result is a variogram model with a nugget of 46.8 cm$^2$, a partial sill of 157 cm$^2$, and a range on the sill of 129 km. We plot the variogram model on top of the empirical variogram by typing
```{r}
plot(MW.v$dist/1000, MW.v$gamma, xlim = c(0, 130), ylim = c(0, 150),
   xlab = "Lag distance (h) [km]",
   ylab = expression(paste("Semivariance (", gamma, ") [", m^2/s^2, "]")),
   las = 1, pch = 19)
grid()
nug <- MW.vm$psill[1]
ps <- MW.vm$psill[2]
r <- MW.vm$range[2] / 1000
h <- seq(0, 130, .2)
fit <- ps * (1 - exp(-h^2 / (r^2))) + nug
lines(h, fit, lwd=2)
```

Let $r$ be the range, $c$ the partial sill and $c_o$ the nugget, then the equation defining the curve over the set of lag distances $h$ is
$$
\gamma(h)=c\left(1-\exp\left(-\frac{h^2}{r^2}\right)\right)+c_o
$$

Check for anisotropy.
```{r}
plot(variogram(gust ~ Longitude + Latitude, 
               data = MW, 
               alpha = c(0, 45, 90, 135)), 
     xlab = "Lag Distance (m)")
```

### Kriging

The final step is to use the variogram model together with the rainfall values at the observation sites to create an interpolated surface.

Interpolation is done using the `krige()` function. The first argument is the model specification and the second is the data.  Two other arguments are needed. One is the variogram model using the argument name model and the other is a set of locations identifying where the interpolations are to be made.  This is specified with the argument name `newdata`.

Here we interpolate first to locations (point kriging) on a regular grid and then to the county polygons (block kriging). To create a grid of locations within the boundary of Florida type
```{r}
grid <- expand.grid(Longitude = seq(min(MW.df$Longitude), max(MW.df$Longitude), l = 61),
                    Latitude = seq(min(MW.df$Latitude), max(MW.df$Latitude), l = 61))
```

```{r}
grid.sp <- grid
gridded(grid.sp) <- ~ Longitude + Latitude
proj4string(grid.sp) <- CRS(ll)
grid.sp <- spTransform(grid.sp, CRS(lcc))
```

```{r}
tm_shape(MW.spdf) +
  tm_bubbles(size = "gust") +
tm_shape(counties_sf) +
  tm_borders() +
tm_shape(grid.sp) +
  tm_dots()
```

First we use the `krige()` function to interpolate  the observed rainfall to the grid locations.For a given location, the interpolation is a weighted average of the rainfall across the entire region where the weights are determined by the variogram model.
```{r}
ipl <- krige(gust ~ Longitude + Latitude, 
             locations = MW.spdf, 
             newdata = grid.sp, 
             model = MW.vm)
```

If the variogram model is not included then inverse distance weighted interpolation is performed. The function will not work if different values share the same location. 

```{r}
tm_shape(ipl) +
  tm_bubbles(size = "gust") +
tm_shape(counties_sf) +
  tm_borders() +
tm_shape(grid.sp) +
  tm_dots()
```


The saved object (`ipl`) inherits the spatial points object specified in the `newdata` argument, but extends it to a spatial data frame by adding a data slot. The data slot is a data frame with two variables.  The first `var1.pred` is the interpolated rainfall and the second `var1.var` is the prediction variance.

We plot the interpolated field using the `spplot` method. We specify an inverted topographic color ramp to highlight in blue regions with the highest rain totals.
```{r}
spplot(ipl, "var1.pred", 
       col.regions = rev(topo.colors(20)))
```
